Supporting Information : Asymmetric tail-dependence in extreme climatic events affects population-extinction risk

### 2019-08-28 MainText.Rmd
methods edited, started to edit intro

<!--how common are the extreme events?-->
For past few decades, there have been increasing evidences that anthropogenic climate change 
is causing extreme events, e.g. temperature or precipitation extremes,
heat-waves, wild-fire etc. [@ummenhofer2017extreme]. Frequent extreme events 
can shift temperature or precipitation-mean, 
change variability of the distribution, as well as change its 
symmetry or skewness [@hansen2012perception]. Such extreme climatic events (ECE) can pose severe
impacts at the individual, population, community and ecosystem level [@bragazza2008climatic; 
@gutschick2003extreme; @jiguet2011community; @mckechnie2009climate;
@hoover2014resistance]. Therefore, 
understanding the long-term impacts of changes 
in ECEs has been a longstanding interest in ecological research 
[@katz2005statistics; @smith2011ecological; @diez2012will; @moritz2013future; @bailey2016tackling].

<!--How extinction risk is related to extreme events : past studies
consequences of extreme events on population life history, community-->
Earlier studies showed ECEs may affect species abundances (species re-ordering) through 
life-history traits and this alteration in community structures could cascade 
through higher hierarchical levels and lead to local extinction of 
species [@thomas2004extinction; @edgar2010nino; @moreno2011extreme]. 
<!--Should I mention idea of match-mismatch hypothesis?-->
For example, 
a study based on annal survival data showed changes
in prey communities induced by extreme flooding affected predators in 
aquatic and adjacent riparian habitats [@chiu2013impact] via trophic linkages. Similarly, 
$43$ years of ring-recovery data was used to examine the demographic impact of ECEs 
through temporal variability in juvenile, immature and adult survival
of European shags on the Isle of May in eastern Scotland [@frederiksen2008demographic].
A recent study [@mcdermott2017sensitivity] investigated the sensitivity 
of $41$ UK butterfly species to four different ECEs (extreme heat, extreme cold, 
extreme precipitation, drought) over $37$ years (1976-2012) at the site level, across 
each species' life-stages. Experimental study showed successive ECEs with short time intervals 
could weaken the resistance of plant communities too [@dreesen2014successive]. 
All these studies indicated that environmental stochasticity (through ECEs) 
can lead to demographic stochasticity (through influence on reproductive or 
mortality rate) of different age class of a population making it vulnerable 
to local extinction.

<!--How copula could be relevent in estimating extreme event's 
influence on extinction risk : introduce Taildep. concept-->
A recent study [@Ghosh_copula] showed the importance, causes 
and consequences of copula structures in associations between ecologically
important variables and demonstrated a basic example where 
copula was used to model extinction risk of a density-independent 
metapopulation. Present manuscript is an extension for the application of copula 
in estimating the extinction risk as introduced by [@Ghosh_copula]. We will show 
why the potential use of copula
in modelling extinction risk goes beyond the density-independent model 
and even applicable for a simple age-structured population where its vital rates are influenced by ECEs. 
Following we will introduce the concept of copula in a manner that follows @Ghosh_copula.

@@

<!--calculation of ext risk with age-structured model-->
To calculate extinction risk after each time step, we simulated model (\ref{eq_age_str}) for $25$ years
with $10000$ simulations. For each simulation, we started with initial population $(E,A)_{t=0} = (5,10)$
and following Eq. \ref{eq_age_str} we checked whether after each time step the total population $(E+A)_{t}$
was below a certain extinction threshold (we set it as $2$) or not. If total population was below that threshold then we 
considered population went extinct for that simulation. We also calculated $2.5^{th}$ and $97.5^{th}$ quantiles
of $10000$ simulations with the probability of computed extinction risk at each time step (for details see [`code`](https://github.com/sghosh89/ERC/blob/master/ext_risk_cop_age_str.R)).

<!--density-dependence metapopulation model-->
We explored (Q2) by examining how nature of the tail dependence of an
environmental variable across space could influence the extinction risk of a 
metapopulation model with density-dependence feedback. We considered six density-dependent 
population-growth models (Ricker, Hassell, Maynard Smith, Pennycuick, Verhulst, Austin-Brewer)
in noisy environment (see Table \ref{tab_model_summary}) 
from a metapopulation context.
Models were adapted from earlier work
by @cohen1995unexpected, excluding only two : Varley model (this is not a continuous one) 
and Malthus-Condorcet-Mill model (terms containing logarithm of extinct population does not make sense). 
All models assumed a generalized form : $\vec{P}_{t+1} = D \lambda_t \vec{P_t}$,
where $\vec{P_t}$ was a vector of population densities, 
the $i^{th}$ component $P_t^i$ representing
the density in the $i^{th}$ habitat patch ($i=1,2,\ldots,n$) at time $t$. 
The $n \times n$ matrix $\lambda(t)$ was diagonal with $i^{th}$ diagonal entry 
of stochastic density dependence form $\lambda_i(t)$ (see second column 
of Table \ref{tab_model_summary}). For simplicity, 
we considered $n$ patches in each model were arranged, evenly spaced, in a line
with the provision of "local" or "global" dispersal among patches. A n $\times$ n dispersal
matrix, $D$ assumed that a fraction $d$ of 
each population patch dispersed during any given time step, equally distributed to
the two or one nearest neighbors in case of local dispersal. For global dispersal, the fraction
$d$ that dispersed from each patch was divided up equally among all other patches. We called $d$ as dispersal
rate with $d$ being zero there was no dispersal, and threby no chance to rescue an extinct 
patch through migration from nearby metapopulation patches.

The environmental noises from any two out of $n$ patches, say, 
$(\epsilon_t^i, \epsilon_t^j)$ were normally 
distributed within each patch with mean $m=0$ and standard deviation $\sigma$ 
(see the specified value $\sigma$ for each model in 3$^{rd}$ 
column of Table \ref{tab_model_summary}). Noises were 
independent through time with same 
spatial correlations between patches (to within sampling error) for every simulation, 
but moderately (or extremely) left- and right-tail dependent between patches in
different simulations.

<!--Shya: ends here-->

A three dimensional array, $n_s \times n \times n_t$, 
$\epsilon$, was made from these extreme tail-dependence noise 
matrix generated via algorithm stated in 
Appendix \ref{SM-extr_tail_dep} and used to calculate extinction risk 
for each of the specified model in Table \ref{tab_model_summary} for $numsteps = 25$ years, $n =$ 5 patches
and $numsims = 10000$ simulations.

The models stated in Table \ref{tab_model_summary} had one thing in 
common that they showed undercompensatory
dynamics (monotonic approach towards stable steady state $P_e$) when
growth parameter $r$ remained below a critical 
level, say, $r_c$. If $r$ was beyond $r_c$, model showed 
overcompensatory dynamics with oscillatory approach 
towards stable equilibrium $P_e$. But if $r$ became too high and went
beyond $r_{bf}$, the parameter value responsible for Hopf-bifurcation,
model equilibrium $P_e$ became unstable and model could show sustained
oscillation and eventually led to chaos. Appendix \ref{SM-model_stability}
summarizes the stability analysis of each model in their deterministic form.

<!--say how you set ext. threshold = Pe/10-->
We started with $n$ patches of initial population $P_{t=0}^i = P_e^i, (i = 1,2,\ldots,n)$ where 
$P_e^i$ was the equilibrium value (or similar to carrying capacity concept) 
for each density-dependence model (see 4$^{th}$ column of Table \ref{tab_model_summary}).
We assumed population went extinct after each time step if it was below a certain threshold.
That extinction threshold was set to 10\% of initial population $P_e$ for individual model.

<!--test hypothesis from Q2 : ext risk with density-dependence metapopulation model with r from different regime-->
To test our hypothesis from (Q2) of \nameref{Introduction} for each metapopulation model,
we estimated extinction risk for a wide range of parameter $r$, 
which covered the undercompensatory and overcompensatory regime. We 
chose equispaced values of $r$ from the 95\% quantiles 
of possible range ($r_{min},r_{bf}$), where $r_{min}$ was the minimum value of $r$ 
for which the model equilibrium $P_e$ had always a finite, positive initial 
population (for details, see Appendix \ref{SM-model_stability}). For each chosen $r$ from each 
model where metapopulation could disperse locally with varying dispersal 
rate $d= 0,0.1,0.2,..0.5$, we calculated extinction risk with left 
tail dependent and right tail dependent environmental noise 
in different simulations, respectively. We also repeated the analysis 
for global dispersal for each metapopulation model. Script 
[`mydispersal.R`](https://github.com/sghosh89/ERC/blob/master/mydispersal.R) was used for this 
analysis.

<!--test hypothesis from Q3 : switching of ext risk with density-dependence metapopulation model with d=0-->
Given hypothesis from (Q2) was valid which scrutinized that whereas 
left tail dependence in noise made extinction risk 
higher in undercompensatory regime ($r < r_c$), it was the 
right tail dependent noise
which appeared to evoke higher extinction risk for metapopulations
in overcompensatory regime ($r > r_c$), we tested our third hypothesis for (Q3).
We considered
the switching to take place near $r_c$, not exactly at $r = r_c$ because the transition point $r_c$
from under to overcompensatory dynamics was expected to differ in stochastic environment than its 
deterministic analog depending on the underlying nonlinearity of the model.
We generated plots with extinction risk along y-axis
for both left tail and right tail dependent noises 
against the varying range of $r$ (along x-axis) when no dispersal ($d=0$) was considered.
We excluded the possibility of dispersal in this case to avoid any rescue effect among patches. 
For each model with extreme left and extreme right tail dependent noises, we calculated 
2.5$^{th}$ and 97.5$^{th}$ quantiles
of 10000 simulations with the probability of computed
extinction risk at each $r$. We compared $r_c$ and the value of $r$ at which 
95\% CI of extinction risk from both noises overlapped
with each other. This could give us an idea for supporting our hypothesis.

@@

 <!--FIG : age structure model results : linear, nonlinear-->
\begin{figure*}[!h]
\textbf{ \hspace{5.4 cm} (A) \hspace{4 cm} (B)} \\
\vspace{-0.9 cm}
\begin{center}
\includegraphics[width=5 cm]{./Results/age_str_results/extrisk_age_str_extremetail_linear.pdf}
\includegraphics[width=5 cm]{./Results/age_str_results/extrisk_age_str_extremetail_nonlinear_omg_100.pdf}\\
\caption{Extinction risk from single patch age structured $(A)$ linear ($\Omega = \infty$) and $(B)$ nonlinear ($\Omega = 100$) models against time (shading represents 95\% confidence interval). Fecundity of adults and survival of eggs have either extreme left or extreme right tail dependence, numsims=10000. \label{fig_extrisk_agestr_etd}}
\end{center}
\end{figure*}

<!--FIG :  summary : all nonlinear model results for local dispersion : extrisk vs. d-->
\begin{figure*}[!h]
\begin{center}
\includegraphics[width=15 cm]{./Results/allmodel_ext_risk_local_disp_extremenoise.pdf}
\caption{For each of six nonlinear density-dependence models presented in Table \ref{tab_model_summary} with local dispersal
among $n=5$ patches, extinction risk after $numsteps = 25$ years for $numsims = 10000$ simulations were calculated.
Extinction risk varies as dispersion-rate ($d$) goes from $0$ to $0.5$. Last two columns represent extinction risk calculated for intrinsic growth rate $(r)$ taken from under-compensatory and over-compensatory regime of deterministic model analog, respectively.\label{fig_extrisk_allmodel_localdisp}}
\end{center}
\end{figure*}

>>>>

<!--how common are the extreme events?-->
For past few decades, extreme climatic events (ECE) appeared more common than 
usual due to anthropogenic climate change, e.g. temperature or precipitation extremes,
heat-waves, wild-fire etc. [@ummenhofer2017extreme]. Frequent extreme events 
can shift temperature or precipitation-mean, 
change variability of the distribution, as well as change its 
symmetry or skewness [@hansen2012perception]. Extreme climatic events can pose severe
impacts at the individual, population, community and ecosystem level [@bragazza2008climatic; 
@gutschick2003extreme; @jiguet2011community; @mckechnie2009climate;
@hoover2014resistance]. Therefore, 
understanding the long-term impacts of changes 
in ECEs has been a longstanding interest in ecological research 
[@katz2005statistics; @smith2011ecological; @diez2012will; @moritz2013future; @bailey2016tackling].

<!--How extinction risk is related to extreme events : past studies
consequences of extreme events on population life history, community-->
Earlier studies showed ECEs may affect species abundances (species re-ordering) through 
life-history traits, alter community structures through cascading effect on 
higher hierarchical levels and lead to local extinction of 
species [@thomas2004extinction; @edgar2010nino; @moreno2011extreme]. 
<!--Should I mention idea of match-mismatch hypothesis?-->
For example, a study [@chiu2013impact] showed extreme flood affected 
the annal survival of a prey community that in turn affected predators in 
aquatic as well as adjacent riparian habitats via trophic linkages. Another study by 
@frederiksen2008demographic used $43$ years of ring-recovery data 
to examine the demographic impact of ECEs 
through temporal variability in juvenile, immature and adult survival
of European shags on the Isle of May in eastern Scotland.
A relatively recent study [@mcdermott2017sensitivity] investigated the sensitivity 
of $41$ UK butterfly species to four different ECEs (extreme heat, extreme cold, 
extreme precipitation, drought) over $37$ years at the site level, across 
each species' life-stages. Experimental study showed successive ECEs with short time intervals 
could weaken the resistance of plant communities too [@dreesen2014successive]. 
All these studies indicated that environmental stochasticity (through ECEs) 
can lead to demographic stochasticity (through influence on reproductive or 
mortality rate) of different age class of a population making it vulnerable 
to local extinction.

<!--Earlier study on metapopulation extinction risk-->

<!--How copula could be relevent in estimating extreme event's 
influence on extinction risk : introduce Taildep. concept-->
A recent study [@Ghosh_copula] used the tail dependence feature 
to estimate extinction risk of a density-independent 
metapopulation model. Present manuscript is an extension for the application of copula 
in estimating the extinction risk as introduced by [@Ghosh_copula]. We will show 
why the potential use of copula in modelling extinction risk goes beyond the density-independent model 
and even applicable for a simple age-structured population where its vital rates are influenced by ECEs. 
Following we will introduce the concept of copula in a manner that follows @Ghosh_copula.

@@

<!--calculation of ext risk with age-structured model-->
To calculate extinction risk after each time step, we simulated model (\ref{eq_age_str}) for $25$ years
with $10000$ simulations. For each simulation, we started with initial population $(E,A)_{t=0} = (5,10)$
and following Eq. \ref{eq_age_str} we checked whether after each time step the total population $(E+A)_{t}$
was below a certain extinction threshold (we set it as $2$) or not. If total population was below that threshold then we 
considered population went extinct for that simulation. We also calculated $2.5^{th}$ and $97.5^{th}$ quantiles
of $10000$ simulations with the probability of computed extinction risk at each time step (see [`code`](https://github.com/sghosh89/ERC/blob/master/ext_risk_cop_age_str.R)).

<!--density-dependence metapopulation model-->
We explored (Q2) by examining how nature of the tail dependence of an
environmental variable across space could influence the extinction risk of a 
metapopulation model with density-dependence feedback. We considered six density-dependent 
population-growth models (Ricker, Hassell, Maynard Smith, Pennycuick, Verhulst, Austin-Brewer)
in noisy environment (see Table \ref{tab_model_summary}) 
from a metapopulation context.
Models were adapted from earlier work
by @cohen1995unexpected, excluding only two : Varley model (this is not a continuous one) 
and Malthus-Condorcet-Mill model (terms containing logarithm of extinct population does not make sense). 
All models assumed a generalized form : $\vec{P}_{t+1} = D \lambda_t \vec{P_t}$,
where $\vec{P_t}$ was a vector of population densities, 
the $i^{th}$ component $P_t^i$ representing
the density in the $i^{th}$ habitat patch ($i=1,2,\ldots,n$) at time $t$. 
The $n \times n$ matrix $\lambda(t)$ was diagonal with $i^{th}$ diagonal entry 
of stochastic density dependence form $\lambda_i(t)$ (see second column 
of Table \ref{tab_model_summary}). For simplicity, 
we considered $n$ patches in each model were arranged, evenly spaced, in a line
with the provision of "local" or "global" dispersal among patches. A n $\times$ n dispersal
matrix, $D$ assumed that a fraction $d$ of 
each population patch dispersed during any given time step, equally distributed to
the two or one nearest neighbors in case of local dispersal. For global dispersal, the fraction
$d$ dispersed equally to other $(n-1)$ patches. We called $d$ as dispersal
rate with $d$ being zero there was no dispersal, and hence no chance to rescue an extinct 
patch through migration from nearby metapopulation patches.

We simulated our metapopulation models with noises $(\epsilon_t^i)$ that were 
independent through time with same spatial correlations between patches (to within sampling error)
for every simulation, but moderately (or extremely) left- and right-tail dependent between patches in
different simulations. For moderately left-tail dependent noise we drew points from a $n$ dimensional 
Clayton copula with same Kendall correlation ($= 0.6$) between any two out of $n$ patches and made 
the marginals normal with zero mean and standard deviation $\sigma$ 
(see the specified value $\sigma$ for each model in 3$^{rd}$ 
column of Table \ref{tab_model_summary}). For moderately 
right-tail dependent noise we flipped ($180^{\circ}$ rotation) the points drawn 
from Clayton copula so that all the statistical properties remained 
same except for the direction of tail-dependence. We generated
extreme left- or right-tail dependent noises via algorithm stated in 
Appendix \ref{SM-extr_tail_dep}. Our models (Table \ref{tab_model_summary}) 
computed extinction risk for $n_t = 25$ years, $n = 5$ patches
and $n_s = 10000$ simulations with both moderate and extreme tail dependence (left-, right-tail).

<!--say how you set ext. threshold = Pe/10-->
We started with $n$ patches of initial population $P_{t=0}^i = P_e^i, (i = 1,2,\ldots,n)$ where 
$P_e^i$ was the equilibrium value (or similar to carrying capacity concept) 
for each density-dependence model (see 4$^{th}$ column of Table \ref{tab_model_summary}).
We assumed population went extinct after each time step if it was below a 
certain threshold (set to 10\% of initial population $P_e$ for individual model).

The models stated in Table \ref{tab_model_summary} had one common 
feature: they showed undercompensatory
dynamics (monotonic approach towards stable steady state $P_e$) when
growth parameter $r$ remained below a critical 
level, say, $r_c$. Models showed 
overcompensatory dynamics with oscillatory approach 
towards stable equilibrium $P_e$ when $r$ was beyond $r_c$. 
If $r$ became too high and exceeded $r_{bf}$ - the parameter value responsible for Hopf-bifurcation -
model equilibrium $P_e$ became unstable. Beyond Hopf-bifurcation point, model showed sustained
oscillation and eventually led to chaos with increasing $r$. Appendix \ref{SM-model_stability}
summarized the stability analysis of each model in their deterministic form.

<!--test hypothesis from Q2 : ext risk with density-dependence metapopulation model with r from different regime-->
To test our hypothesis from (Q2) of \nameref{Introduction} for each metapopulation model,
we estimated extinction risk for a wide range of parameter $r$ 
that covered the undercompensatory and overcompensatory regime. We 
chose equispaced values of $r$ from the 95\% quantiles 
of possible range ($r_{min},r_{bf}$), where $r_{min}$ was the minimum value of $r$ 
to get a positive, finite equilibrium $P_e$ (for details, see Appendix 
\ref{SM-model_stability}) to start with ($t=0$). For each chosen $r$,
we calculated extinction risk after $25$ years with each metapopulation model 
where population could disperse locally (or globally) with varying dispersal 
rate $d= 0,0.1,0.2,..0.5$ and extreme (or moderate) left 
tail- and right-tail dependent environmental noise 
in different simulations (for code see [`mydispersal.R`](https://github.com/sghosh89/ERC/blob/master/mydispersal.R)).

<!--test hypothesis from Q3 : switching of ext risk with density-dependence metapopulation model with d=0-->
Given hypothesis from (Q2) was valid - that left tail dependent noise made extinction risk 
higher in undercompensatory regime ($r < r_c$) and
right tail dependent noise made extinction risk higher
in overcompensatory regime ($r > r_c$) - we tested our third hypothesis for (Q3).
We expected the switching to take place near $r_c$, not exactly at $r = r_c$ because the transition point $r_c$
from under to overcompensatory dynamics differed in stochastic environment than its 
deterministic analog depending on the underlying nonlinearity of the model.
We plotted extinction risk along y-axis
for both left tail and right tail dependent noises 
against the varying range of $r$ along x-axis when no dispersal ($d=0$) was considered.
We excluded the possibility of dispersal in this case to avoid any rescue effect among patches. 
For each model with extreme (or moderate) left- and right-tail dependent noises, we calculated 
2.5$^{th}$ and 97.5$^{th}$ quantiles
of 10000 simulations with the probability of computed
extinction risk at each $r$. We compared $r_c$ and the value of $r$ at which 
95\% CI of extinction risk from both noises overlapped
with each other. This could give us an idea for supporting our hypothesis.

@@

 <!--FIG : age structure model results : linear, nonlinear-->
\begin{figure*}[!h]
\textbf{ \hspace{5.4 cm} (A) \hspace{4 cm} (B)} \\
\vspace{-0.9 cm}
\begin{center}
\includegraphics[width=5 cm]{./Results/age_str_results/extrisk_age_str_extremetail_linear.pdf}
\includegraphics[width=5 cm]{./Results/age_str_results/extrisk_age_str_extremetail_nonlinear_omg_100.pdf}\\
\caption{Extinction risk from single patch age structured $(A)$ linear ($\Omega = \infty$) and $(B)$ nonlinear ($\Omega = 100$) models against time (shading represents 95\% confidence interval). Fecundity of adults and survival of eggs have either extreme left or extreme right tail dependence, $n_s=10000$. \label{fig_extrisk_agestr_etd}}
\end{center}
\end{figure*}

<!--FIG :  summary : all nonlinear model results for local dispersion : extrisk vs. d-->
\begin{figure*}[!h]
\begin{center}
\includegraphics[width=15 cm]{./Results/allmodel_ext_risk_local_disp_extremenoise.pdf}
\caption{For each of six nonlinear density-dependence models presented in Table \ref{tab_model_summary} with local dispersal
among $n=5$ patches, extinction risk after $n_t = 25$ years for $n_s = 10000$ simulations were calculated.
Extinction risk varies as dispersion-rate ($d$) goes from $0$ to $0.5$. Last two columns represent extinction risk calculated for intrinsic growth rate $(r)$ taken from under-compensatory and over-compensatory regime of deterministic model analog, respectively.\label{fig_extrisk_allmodel_localdisp}}
\end{center}
\end{figure*}

### 2019-08-28 SuppMat.Rmd
editing methods...continued
methods edited, started to edit intro

For extreme tail-dependence algorithm, a $N \times n$ matrix, M was generated of independent 
standard-normal random draws (in general), where $N = numsteps(=25) \times numsims(=10000)$.
Then, each row of
M was identified for a first set of modifications
randomly and independently with a 50\% chance; for those rows identified, every
element of the row was replaced with the absolute
value of the first element of the row (including the first element itself).
For other rows, every element in the row was replaced by the negative of 
its own absolute value. This produces right-tail dependent
matrix, normally distributed for each column. If left-tail 
dependent noise is desired, take the negative of the entire matrix.
Our [code](https://github.com/sghosh89/ERC/blob/master/ExtremeTailDep.R) for
generating extreme tail dependence matrix is flexible enough, so that 
matrix M can be generated initially with independent random draws from a normal 
distribution with a specified mean and standard deviation.

>>>>

For extreme tail-dependence algorithm, a $N \times n$ matrix, M was generated of independent 
standard-normal random draws (in general). <!--where $N = numsteps(=25) \times numsims(=10000)$ : only for 2d case-->.
Then, each row of M was identified for a first set of modifications
randomly and independently with a 50\% chance; for those rows identified, every
element of the row was replaced with the absolute
value of the first element of the row (including the first element itself).
For other rows, every element in the row was replaced by the negative of 
its own absolute value. This produces right-tail dependent
matrix, normally distributed for each column. If left-tail 
dependent noise is desired, take the negative of the entire matrix.
Our [code](https://github.com/sghosh89/ERC/blob/master/ExtremeTailDep.R) for
generating extreme tail dependence matrix is flexible enough, so that 
matrix M can be generated initially with independent random draws from a normal 
distribution with a specified mean and standard deviation.
